---
title: "Cyslistic Project"
author: "Nkechi Ihewulezi"
date: "2022-09-10"
output:
  html_document: default
  pdf_document: default
---


The purpose of this script is to consolidate downloaded 2021 Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: “In what ways do members and casual riders use Divvy bikes differently?”


```{r load Relevant Library}
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data
rm(list = ls())

```

## STEP 1: COLLECT DATA

```{r Upload Divvy datasets (csv files) here}
df1 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202101-divvy-tripdata.csv")
df2 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202102-divvy-tripdata.csv")
df3 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202103-divvy-tripdata.csv")
df4 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202104-divvy-tripdata.csv")
df5 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202105-divvy-tripdata.csv")
df6 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202106-divvy-tripdata.csv")
df7 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202107-divvy-tripdata.csv")
df8 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202108-divvy-tripdata.csv")
df9 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202109-divvy-tripdata.csv")
df10 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202110-divvy-tripdata.csv")
df11 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202111-divvy-tripdata.csv")
df12 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202112-divvy-tripdata.csv")

```

## STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE

```{r  Compare column names each of the files}
# While the names don't have to be in the same order, they DO need to match perfectly before we can use a command to join them into one file
colnames(df1)
colnames(df2)
colnames(df3)
colnames(df4)
colnames(df5)
colnames(df6)
colnames(df7)
colnames(df8)
colnames(df9)
colnames(df10)
colnames(df11)
colnames(df12)

```

Since everything is in order, we proceed 
```{r Stack individual data frames into one big data frame}
biketrips_2021  <- bind_rows(df1,df2,df3,df4,df5,df6,df7,df8,df9,df10,df11,df12)
# Inspect the data frame and look for incongruencies

str(biketrips_2021)

```
All data structures are in order except the start and end date that are in character format. This has to be corrected to date and time format.
```{r Changing to suitable data type}
biketrips_2021$started_at = lubridate::dmy_hm(biketrips_2021$started_at)

biketrips_2021$ended_at = lubridate::dmy_hm(biketrips_2021$ended_at)
```
# STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS

```{r}
# Inspect the new table that has been created
colnames(biketrips_2021)  #List of column names
nrow(biketrips_2021)  #How many rows are in data frame?
dim(biketrips_2021)  #Dimensions of the data frame?
head(biketrips_2021)  #See the first 6 rows of data frame.  Also 
tail(biketrips_2021) # See the last 6 rows of data frame
str(biketrips_2021)  #See list of columns and data types (numeric, character, etc)

```
```{r}
# There are a few problems we will need to fix:
# (1)The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.
# (2) We will want to add a calculated field for length of ride since the 2021 data does not have the "trip duration" column. We will add "ride_length" to the entire dataframe for trip duration.
# (3) There are some rides where trip duration shows up as negative and zero. We will want to delete these rides.

# (4) Add columns that list the date, month, day, and year of each ride
# This will allow us to aggregate ride data for each month, day, or day of the week ... before completing these operations we could only aggregate at the ride level
biketrips_2021$start_date <- as.Date(biketrips_2021$started_at) #The default format is yyyy-mm-dd
biketrips_2021$month <- format(as.Date(biketrips_2021$start_date), "%m")
biketrips_2021$day <- format(as.Date(biketrips_2021$start_date), "%d")
biketrips_2021$day_of_week <- format(as.Date(biketrips_2021$start_date), "%A")

biketrips_2021$start_hour <- lubridate::hour(biketrips_2021$started_at)
biketrips_2021$end_hour <- lubridate::hour(biketrips_2021$ended_at)


# Add a "ride_length" calculation to all_trips (in seconds)

biketrips_2021$ride_length <- difftime(biketrips_2021$ended_at,biketrips_2021$started_at, units =c("sec"))

# Inspect the structure of the columns
str(biketrips_2021)

# Convert "ride_length" from Factor to numeric so we can run calculations on the data
is.factor(biketrips_2021$ride_length)
biketrips_2021$ride_length <- as.numeric(as.character(biketrips_2021$ride_length))
is.numeric(biketrips_2021$ride_length)

# Remove "bad" data
# The data frame includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative
# We will create a new version of the data frame since data is being removed
# remove empty rows/cols.
biketrips_2021 <- janitor::remove_empty(biketrips_2021, which = c("cols"))
biketrips_2021 <- janitor::remove_empty(biketrips_2021, which = c("rows"))
trips_2021 <- biketrips_2021 %>% filter(ride_length > 0) %>% drop_na() %>%   select(-ride_id, -start_lat,-start_lng,-end_lat,end_lng)

```
## STEP 4: CONDUCT DESCRIPTIVE ANALYSIS
```{r}
# Descriptive analysis on ride_length (all figures in seconds)
mean(trips_2021$ride_length) #straight average (total ride length / rides)
median(trips_2021$ride_length) #midpoint number in the ascending array of ride lengths
max(trips_2021$ride_length) #longest ride
min(trips_2021$ride_length) #shortest ride

# You can condense the four lines above to one line using summary() on the specific attribute
summary(trips_2021$ride_length)

# Compare members and casual users
aggregate(trips_2021$ride_length ~ trips_2021$member_casual, FUN = mean)
aggregate(trips_2021$ride_length ~ trips_2021$member_casual, FUN = median)
aggregate(trips_2021$ride_length ~ trips_2021$member_casual, FUN = max)
aggregate(trips_2021$ride_length ~ trips_2021$member_casual, FUN = min)

# See the average ride time by each day for members vs casual users
aggregate(trips_2021$ride_length ~ trips_2021$member_casual + trips_2021$day_of_week, FUN = mean)

# Notice that the days of the week are out of order. Let's fix that.
trips_2021$day_of_week <- ordered(trips_2021$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

# Now, let's run the average ride time by each day for members vs casual users
aggregate(trips_2021$ride_length ~ trips_2021$member_casual + trips_2021$day_of_week, FUN = mean)

# analyze ridership data by type and weekday
trips_2021 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts

# Let's visualize the number of rides by rider type
trips_2021 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")

# Let's create a visualization for average duration
trips_2021 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

# Top Ten (10) Start Stations

  biketrips_2021 %>%  count(start_station_name,SORT = TRUE) %>% top_n(10) %>%
  ggplot() + geom_col(aes(x=reorder(start_station_name,n),y=n)) + coord_flip()
# Bike Trips Per month
 biketrips_2021 %>% count(started_at,sort = TRUE) %>%
  ggplot() + geom_line(aes(x=started_at,n)) +
  geom_smooth(aes(x=started_at,y=n),method = "gam")

```
# STEP 5: EXPORT SUMMARY FILE FOR FURTHER ANALYSIS
```{r}
# Create a csv file that we will visualize in Excel, Tableau or any presentation software of your choice
# create a summary dataframe
bikerides_2021 <- trips_2021 %>%
  group_by(member_casual,rideable_type,start_date,day_of_week) %>%
                summarise(Total_ride_length = sum(ride_length),
                  Average_ride_length = mean(ride_length),
                  Median = median(ride_length),
                  Max_ride_length = max(ride_length),
                  Min_ride_length = min(ride_length),
                  Count = n()
                ) %>% ungroup()
write.csv(bikerides_2021, file = 'C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/bikerides_2021.csv')

# https://public.tableau.com/app/profile/nkechi.ihewulezi/viz/GoogleCapstonePresentation/Cyclistic?publish=yes

```

