---
title: "Google_Capstone_Project"
author: "Nkechi Ihewulezi"
date: "2022-09-14"
output:
  pdf_document: default
  html_document: default
---
## Case Study: How Does a Bike-Share Navigate Speedy Success?
**Scenario**
You are a junior data analyst working in the marketing analyst team at Cyclistic, a bike-share company in Chicago. The director
of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore,
your team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights,
your team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives
must approve your recommendations, so they must be backed up with compelling data insights and professional data
visualizations.

## Step 1: Load packages
Start by installing the required packages. If you have already installed and loaded `tidyverse`, `skimr`, and `janitor` in this session, feel free to skip the code chunks in this step. This may take a few minutes to run, and you may get a pop-up window asking if you want to proceed. Click yes to continue installing the packages.
Once a package is installed, you can load it by running the `library()` function with the package name inside the parentheses:
```{r load packages}
library(tidyverse)
library(scales)
library(lubridate)
library(skimr)
library(janitor)
rm(list = ls())
```
## Load the data into R
Import previous 12 months data. In this case data for year 2021
Get the file list in the directory and combine into one data frame.


```{r loading the csv files from the dir}
df1 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202101-divvy-tripdata.csv")
df2 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202102-divvy-tripdata.csv")
df3 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202103-divvy-tripdata.csv")
df4 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202104-divvy-tripdata.csv")
df5 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202105-divvy-tripdata.csv")
df6 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202106-divvy-tripdata.csv")
df7 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202107-divvy-tripdata.csv")
df8 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202108-divvy-tripdata.csv")
df9 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202109-divvy-tripdata.csv")
df10 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202110-divvy-tripdata.csv")
df11 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202111-divvy-tripdata.csv")
df12 <- read.csv("C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers/Bikers_data_2021/202112-divvy-tripdata.csv")
```
<!-- Combine the 12 data.frames into one(1) data.frame -->
The files are of same number of variables and as such we can combine into one.

```{r combine with rbind}  
# combine the 12 data frames into one data frame
bike_trips_2021 <- rbind(df1,df2,df3,df4,df5,df6,df7,df8,df9,df10,df11,df12)   
# remove empty rows/cols.
bike_trips_2021 <- janitor::remove_empty(bike_trips_2021, which = c("cols"))
bike_trips_2021 <- janitor::remove_empty(bike_trips_2021, which = c("rows"))
```
## Data cleaning
check data structures to be sure they are in order
```{r }
str(bike_trips_2021)
```

## Changing to suitable data types
All data structures are in order except the start and end date that are in character format. This has to be corrected to date and time format.
```{r}
bike_trips_2021$started_at = lubridate::dmy_hm(bike_trips_2021$started_at)    
bike_trips_2021$ended_at = lubridate::dmy_hm(bike_trips_2021$ended_at)
# confirm Data structure
str(bike_trips_2021)
```
## Data Manipulation
Adding calculated field(trip_duration), extracting start_date, end_date , start_time and end_time and weekday column.

```{r Manipulation of data}

bike_trips_2021$ride_length <- difftime(bike_trips_2021$ended_at, bike_trips_2021$started_at, units = c("mins")) 

bike_trips_2021$start_date <- as.Date(bike_trips_2021$started_at)
bike_trips_2021$end_date <- as.Date(bike_trips_2021$ended_at)

bike_trips_2021$start_hour <- lubridate::hour(bike_trips_2021$started_at)
bike_trips_2021$end_hour <- lubridate::hour(bike_trips_2021$ended_at)


bike_trips_2021$day_of_week <- wday(bike_trips_2021$started_at, label = TRUE)
#drop rows with ride_length <= 0 and columns not needed
df <- bike_trips_2021 %>% filter(ride_length > 0) %>% drop_na() %>% 
  select(-ride_id, -end_station_name, -end_station_id)
bike_trips_2021 <- df

```
## view Cleaned Data

```{r}
head(bike_trips_2021)
```
 
```{r arrange }
#**Arrange Data via ride_length to see the min ride_length**
head(arrange(bike_trips_2021, (bike_trips_2021$ride_length)))

```
##Create summary data frame
```{r}
bike_trips_2021_summ <- bike_trips_2021 %>%
  group_by(member_casual,rideable_type,start_date,day_of_week) %>%
                summarise(Total_ride_length = sum(ride_length),
                  Average_ride_length = mean(ride_length),
                  Median = median(ride_length),
                  Max_ride_length = max(ride_length),
                  Min_ride_length = min(ride_length),
                  Count = n()
                ) %>% ungroup()
```


```{r}
head(bike_trips_2021_summ)

```
## Top Ten (10) Start Stations
```{r}
  bike_trips_2021 %>%  count(start_station_name,SORT = TRUE) %>% top_n(10) %>%
  ggplot() + geom_col(aes(x=reorder(start_station_name,n),y=n)) + coord_flip()
```
 
```{r  Bike Trips Per month} 
#**Bike Trips Per month**
 bike_trips_2021 %>% count(start_date,sort = TRUE) %>%
  ggplot() + geom_line(aes(x=start_date,n)) +
  geom_smooth(aes(x=start_date,y=n),method = "gam")

```
```{r}
#**Count by Bike Trips per day_of_Week**
ggplot(bike_trips_2021_summ) + geom_col(aes(x=day_of_week,y=Count,fill=member_casual)) +
  scale_y_continuous(labels = comma) +
  labs(title="Count of Rides per weekday")

```
```{r}
bike_trips_2021_summ %>% ggplot() + geom_col(aes(x=start_date,y=Count, color=member_casual)) +
  scale_y_continuous(labels = comma) +
  labs(title = "Average of Rides per Day",
       subtitle = "(Bases on 28 day moving average",
       y="Average rides per day") 

```
```{r}
#**Export clean Data**
write.csv(bike_trips_2021_summ,"C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers\\bike_trips_2021_summ.csv",  row.names = FALSE)
write.csv(bike_trips_2021,"C:/Users/Nkechi Ihewulezi/Downloads/R/Bikers\\bike_trips_2021.csv",  row.names = FALSE)
```
 
 
 